{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-10-08 thesis/landslide.tex
landslide - heuristic loop/sync/deadlock

Therefore, Landslide's ICB implementation combines with DPOR
when tagging new branches to explore at the end of each branch:
just as DPOR skips alternate interleavings that are memory-independent,
ICB further filters interleavings requiring more preemptions than the current bound
out of the to-explore set.
The macro {\tt ICB\_BLOCKED},
defined in {\tt schedule.h},
decides if a given thread would require a preemption beyond the current bound to switch to%
\footnote{Since ``voluntary'' context switches (e.g. arising from {\tt yield()})
are often necessary for correct execution,
{\tt ICB\_BLOCKED} does not count such switches towards the preemption count.
Therefore, within a certain preemption bound $B$,
interleavings with more than $B$ context switches may still be tested.
}.
The DPOR implementation then checks, for some $I_{ij}$ it wants to mark for exploration,
whether {\tt ICB\_BLOCKED(}$T_j${\tt )} at the state after $t_i$,
and skips it if so ({\tt tag\_good\_sibling()}/{\tt tag\_all\_siblings()}).

@@

\subsection{Heuristic loop, deadlock, and synchronization detection}
\label{sec:landslide-blocking}

% TODO

>>>>

Therefore, Landslide's ICB implementation combines with DPOR
when tagging new branches to explore at the end of each branch:
just as DPOR skips alternate interleavings that are memory-independent,
ICB further filters interleavings requiring more preemptions than the current bound
out of the to-explore set.
The macro {\tt ICB\_BLOCKED},
defined in {\tt schedule.h},
decides if a given thread would require a preemption beyond the current bound to switch to%
\footnote{Since ``voluntary'' context switches (e.g. arising from {\tt yield()})
are often necessary for correct execution,
{\tt ICB\_BLOCKED} does not count such switches towards the preemption count.
Therefore, within a certain preemption bound $B$,
interleavings with more than $B$ context switches may still be tested.
}.
The DPOR implementation then checks, for some $I_{ij}$ it wants to mark for exploration,
whether {\tt ICB\_BLOCKED}($T_j$) at the state after $t_i$,
and skips it if so ({\tt tag\_good\_sibling()}/{\tt tag\_all\_siblings()}).

@@

\subsection{Heuristic loop, synchronization, and deadlock detection}
\label{sec:landslide-blocking}

Despite the ease of automatically instrumenting a fixed concurrency API such as P2's,
the variety of student implementations inevitably results in many behaviours outside Landslide's model.
This section documents the heuristics Landslide uses to approximate a program's formal behaviour in such situations.

\subsubsection{Infinite loop detection}

All modern presentations of stateless model checking assume finite program length.
Even though all test cases used in this thesis's experiments are hand-written to ensure runtime
not just finite but also short (on account of exponential state space sizes),
bugs may still cause a program to get stuck in an infinite loop unexpectedly.
Detecting such loops in general is of course uncomputable \cite{entscheidungsproblem},
but Landslide has the benefit of knowledge from past iterations to inform its sense of how long the program ``should'' run.

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{cc}
			\includegraphics[width=0.28\textwidth]{heuristic-livelock.pdf}
			&
			\includegraphics[width=0.28\textwidth]{heuristic-tightloop.pdf}
			\\
			(a) Infinite loop around preemption points.
			&
			(b) Stuck between preemption points.
		\end{tabular}
	\end{center}
	\caption{Detecting infinite loops heuristically by comparison to past interleavings.}
	\label{fig:heuristic-loop}
\end{figure}

Landslide checks for two distinct categories of infinite loops,
visualized in Figure~\ref{fig:heuristic-loop}.
Firstly, it keeps a running average of how many preemption points deep each completed interleaving has been in the past.
Whenever a new preemption point is reached, Landslide checks
if its depth is greater than the heuristic constant factor of 20
({\tt PROGRESS\_DEPTH\_\allowbreak{}FACTOR})
times the previous average.
If the depth exceeds that cutoff, it reports an infinite loop bug.
Whether this represents a livelock or just a mundane sequential logic bug is for the user to decide.
However, as
different interleavings do
often execute different program logic and vary in length accordingly,
Landslide waits to test 10 interleavings ({\tt PROGRESS\_CONFIDENT\_\allowbreak{}BRANCHES})
before applying this heuristic;
in the case of fewer,
it scales the depth factor by a heuristic exponential factor of 1.1
to represent its lower confidence
({\tt PROGRESS\_BRANCH\_UNCERTAINTY\_EXPONENT}).
In the special case of the first branch ever tested,
Landslide will abort after a fixed preemption point depth limit of 4000 ({\tt TOO\_DEEP\_0TH\_BRANCH})
-- a program with such $N$ would probably have an impossibly large state space anyway.

Secondly, a program may get stuck in a loop where no preemption points are encountered each new iteration.
For example, race-induced data corruption may cause a list to end up circularly linked, %oops
leading a search or append operation to fail to terminate.
Landslide also maintains a running average of instructions per transition,
and each instruction, compares if more instructions have elapsed since the last preemption point
than a constant times that average.
Because transitions may themselves vary greatly in number of instructions as each represents completely different program logic,
this heuristic is much more lenient than the preemption-point-counting one,
using a multiplicative factor of 4000 ({\tt PROGRESS\_TRIGGER\_FACTOR}).
If such a loop is encountered within the P2 synchronization primitives,
which should generally be free of $O(n)$ operations,
Landslide uses a more aggressive cutoff of 2000
({\tt PROGRESS\_AGGRESSIVE\_TRIGGER\_FACTOR}).
Also, as transitions between data-race preemption points may have as few as 1 instruction each,
Landslide caps the average from below at a minimum of 1000 ({\tt PROGRESS\_MIN\_TRIGGER\_AVERAGE})
to keep the average relatively stable.
%(100 was found, empirically, to be not enough).

These checks are both implemented in {\tt check\_infinite\_loop()} in {\tt landslide.c}.

\subsubsection{Yield-loop detection}

One very common student implementation pattern in P2s and kernels is to open-code
synchronization between threads using a loop that spins around a condition that the loop itself cannot fulfill,
waiting for another thread to allow it to proceed,
rather than using the established synchronization API%
\footnote{Even when the student doesn't open-code any such synchronization and always uses {\tt deschedule}
or primitives built thereupon,
{\tt mutex\_test} (\sect{\ref{sec:education-pebbles-tests}})
often relies on this functionality, as data-race preemption points within {\tt mutex\_lock()}
would otherwise disrupt Landslide's ability to recognize blocking on a contended lock.}.
If Landslide failed to recognize that the thread was in principle blocked,
just as if it had called {\tt deschedule} or {\tt cond\_wait()},
this would result in an infinitely deep branch
as it keeps trying to schedule the waiting thread,
blocking all of its attempted context switches to the thread that could make progress.
These loops need not necessarily {\tt yield} each iteration:
they may spin blindly,
assuming progress is being made on another CPU,
which is not possible under Landslide as it serializes execution.

Landslide detects such ad-hoc {\em yield-loop blocking}
by keeping a counter for each thread to track how many {\tt yield}s it has invoked since the last ``interesting'' activity.
Interesting here is defined heuristically as any other known P2 API call,
with the exception of {\tt mutex\_lock()} and {\tt mutex\_unlock()}%
\footnote{Landslide must recognize yield-blocking loops that contain mutex operations
to allow for open-coded reimplementations of {\tt cond\_wait()},
% see comment above "USER_MUTEX_YIELD_ACTIVITY" in user sync.c
which for example the {\tt paraguay} test uses intentionally,
because it is testing the correctness of the student's {\tt cond\_wait()}.}.
Whenever this counter reaches the heuristic cutoff of 10 ({\tt TOO\_MANY\_YIELDS}),
Landslide declares the thread blocked,
and treats it just as if it had invoked {\tt deschedule()} for purposes of DPOR
({\tt check\_user\_yield\_activity()} in {\tt user\_sync.c}).
% this also involves propagating yield-blockedness back up the branch, along those 10 pps where it was still counting up to 10
% and also refusing to run any other threads while yield is going on - KEEP_RUNNING_YIELDING_THREADS
%
In cases where {\tt yield} itself is not involved,
Landslide also counts the number of atomic instructions ({\tt xchg}, {\tt xadd}, {\tt cmpxchg}, etc),
and marks the thread blocked if it exceeds 100 such ({\tt TOO\_MANY\_XCHGS\_TIGHT\_LOOP})%
\footnote{If preemption points exist in between, instead only 20 such ({\tt TOO\_MANY\_XCHGS\_WITH\_PPS}),
to avoid stressing DPOR's $O(n)$ independence computation.}.
When such a loop contains neither {\tt yield} nor atomics,
Landslide falls back on the standard infinite loop detector and reports a bug directly, as described above.
Future work could extend this to include more modern ways of establishing memory safety such as
acquire/release barriers
%\cite{sully-thesis}
and hardware transactions.

In order to detect when a thread should be unblocked from its yield loop,
Landslide simply leverages DPOR's existing computation of memory conflicts:
whatever condition the blocked thread was waiting for
will show up in the conflicts between it and whichever thread fulfills it.
Hence, every memory conflict detected during DPOR is also checked against any
currently yield-loop-blocked threads, unblocking them in the case of a match
({\tt check\_unblock\_yield\_loop()} in {\tt user\_sync.c}).
If that memory access is not sufficient to let the blocked thread start making progress again,
it will simply trigger the yield-blocking heuristic again.
In theory, this could result in a livelock between two threads,
each in principle blocked in a yield loop,
but where the conditions of the loop happen to conflict with each other,
causing the threads to keep waking each other up;
however,
% hmm... is this true? like if they use mutexes on each other or smth, or use HTM to check a condition
% where the htm failure path uses a stop the world thing...
I have never observed this in practice, as the conditions of such blocking loops are usually read-only.
Future work could heuristically address this by deprioritizing such threads,
as measured by the number of times they've yield-blocked,
so that a third thread which could actually make progress may run if it exists.

% mutex learning
% to unblock from open coded accesses in eg unlock_and_vanish
% this is more for blocked_on_addr stuff, not yield blocked stuff
% kind of clunky and uninteresting to talk abt tbh

\subsubsection{False-positive deadlock avoidance}

In cases where the yield-loop heuristic described above produces false positives,
i.e., blocking a thread which could make progress on its own after all,
Landslide must avoid reporting a deadlock bug if no other thread ends up waking it up through memory conflicts.
After all other threads quiesce,
which ordinarily would trigger deadlock detection,
Landslide checks the system for any yield-looping threads that were blocked heuristically%
\footnote{This includes threads blocked on specific mutexes using the {\tt blocked\_on\_addr} field,
which is set when {\tt yield} is invoked within {\tt mutex\_lock()} without waiting for 10 loops.}.
If any exist, it forces them awake and allows them to proceed;
if they are truly blocked in principle,
they will merely trip the yield-loop limit and go back to sleep again.

Two other forms of heuristic blocking exist in Landslide which are also subject to this retry procedure.
Firstly, when an interleaving has already exhausted the number of preemptions allowed under ICB's current bound,
other threads which are runnable but which require preemptions to switch to are considered ``ICB-blocked''.
When yield-looping mixes with ICB-blocking, the yielding thread will require a preemption
for the other thread to fulfill its blocking condition.
In such a case, simply spending all 128 deadlock-avoidance retries on the yield-looping thread will not solve anything,
so the ICB-blocked thread must be forced awake with higher priority,
disregarding the preemption bound,
to allow the system to progress.
Secondly, in programs which use HTM,
threads blocked by retry sets (\sect{\ref{sec:tm-retrysets}})
must be forced awake with priority before yield-blocked threads,
along similar reasoning.

Landslide will retry this process up to a heuristic limit of 128 times ({\tt DEADLOCK\_FP\_MAX\_\allowbreak()ATTEMPTS}
before issuing a deadlock bug report.
The overhead of this check is linear time in the number of retries permitted,
although this is typically negligible compared to the exponential size of state spaces overall.
If the heuristic limit is too small, very ``loopy'' programs
(which invoke xchg or yield therein)
could falsely exhaust this limit while not being truly deadlocked,
so a good limit should be well higher than the total number of concurrency events expected for any Landslide-friendly test.
The cost of a high limit manifests in the length of preemption traces when deadlock is declared,
which will display a proportional number of meaningless preemption points,
although future work could easily truncate them retrospectively.
This check is implemented in {\tt try\_avoid\_fp\_deadlock()} in {\tt arbiter.c}.
